Hijacking Malaria Simulators with Probabilistic Programming

### 2019-04-26 paper/icml_ai4good/main.tex
updated

\begin{abstract}
Epidemiology simulations have become a fundamental tool in the fight
against the epidemics of various infectious diseases like AIDS and
malaria.
However, the complicated and stochastic nature of these simulators
can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models. This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS). In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks. 
We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators. This reinstates much needed trust between policymakers and evidence-based methods.

@@

To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations CITATIONS. 
These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.

@@

In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

@@

 Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
can be used to express probabilistic models and consequently perform automated inference these. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.

\textit{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost\cite{}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.

The PPS and the simulator exchange xTensor objects \cite{} through TCP or ICP using a generic FlatBuffers \cite{TODO} protocol. On the PPS side, sampling is done using the deep learning framework PyTorch \cite{}.

>>>>

\begin{abstract}
Epidemiology simulations have become a fundamental tool in the fight
against the epidemics of various infectious diseases like AIDS and
malaria.
However, the complicated and stochastic nature of these simulators
can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models. This is achieved by \emph{hijacking} the internal random number generator calls, through the use of a universal probabilistic programming system (PPS). In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks. 
We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators. This reinstates much-needed trust between policymakers and evidence-based methods.

@@

To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus often times relying on computational simulations~\cite{who_global_2015}. 
These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.

@@

In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

@@

 Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
can be used to express probabilistic models and consequently perform automated inference thereon. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{ranganath2014black} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used out-of-the-box by non-experts.

\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.

The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol. On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
